\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-22 paper.tex
save

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likehood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem by minimizing variationial upper bounds of the adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution  matches the observed data.
\glnote{Mention experimental results.}
\glnote{Add 'so what?' conclusion.}

@@

In this note, we develop a likelihood-free inference algorithm for fitting a
forward non-differentiable generative model to observed data. We propose to
adapt the adversarial training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper bounds of the adversarial objectives
respectively with stochastic gradient descent and  natural evolution strategies~\citep{2011arXiv1106.4487W}.
The procedure results in learning a proposal distribution over
simulator parameters, hence producing an abitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

\subsection{Natural evolution strategies}

@@

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound $U(\psi)$ is differentiable, and using the log-likelihood trick it comes:
\begin{align}\label{eqn:approx-grad}
    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] \nonumber \\
    &= \nabla_\psi \int f(\mathbf{c})  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \int f(\mathbf{c}) \nabla_\psi q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \int \left[ f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c}) \right]  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c})]
\end{align}
Effectively, this means that provided that the score function $\nabla_\psi \log q_\psi(\mathbf{c})$ of the proposal
is known and that one can evaluate $f(\mathbf{\mathbf{c}})$ for any $\mathbf{c}$, then
one can construct empirical estimates of Eqn.~\ref{eqn:approx-grad}, which can
in turn be used to perform stochastic gradient descent (or a variant thereof)
in order to minimize $U(\psi)$.

In this simple form, stochastic gradient descent based on
Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
of scale invariance. To avoid the dependence on the parameterization of the
proposal distribution, natural gradients~\citep{amari1998natural} can be used
instead to point towards the steepest descent in the space of realizable
proposal distributions, rather than in the space of their parameters. Effectively,
the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
corrects the gradient of $U$ according to the local curvature of the KL-divergence
surface of proposal distributions, which results in forming
\begin{equation}\label{eqn:approx-grad-nat}
    \tilde{\nabla}_\psi U(\psi) = \mathbf{F}(\psi)^{-1} \nabla_\psi U(\psi),
\end{equation}
where $\mathbf{F}(\psi)$ is the Fisher information matrix
\begin{equation}
    \mathbf{F}(\psi) = \mathbb{E}_{\theta \sim q_\phi}[ \nabla_\psi \log q_\psi(\theta) \nabla_\psi \log q_\psi(\theta)^{\top} ]
\end{equation}
and can be estimated from samples, in particular by reusing the derivatives $\nabla_\psi \log q_\psi(\theta)$ already
evaluated to form $\nabla_\psi U(\psi)$.

@@

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q_\psi \leftarrow \text{prior on $\theta$ (with differentiable and known likelihood)}$}
        \While{$\psi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \}_{m=1}^M \sim p_r$, $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$, $\{\epsilon_m \}_{m=1}^M \sim U[0,1]$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$L^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{RMSProp}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M L^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$.}  \Comment{Update $q_\psi$}
            \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
            \State{$\mathbf{F}(\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi \log q_\psi(\theta_m) \nabla_\psi \log q_\psi(\theta_m)^{\top}$}
            \State{$\nabla_\psi H(q_\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m)$)}
            \State{$\psi \leftarrow \text{SGD}(\mathbf{F}(\psi)^{-1} \nabla_\psi U_g + \gamma \nabla_\psi H(q_\psi)$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

@@

In this work, we propose to rely on evolution strategies to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q_\psi(\theta)$ over the
parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$g$, estimates of $\nabla_\psi U_g$ and $\tilde{\nabla}_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
% That is,
% \begin{equation}
%     \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
% \end{equation} which we can approximate with mini-batches of
% generated data
% \begin{equation}
%     \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
% \end{equation}
% for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
% In practice, using
% \begin{equation}
%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
% \end{equation}
% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}

@@

For completeness, Algorithm~\ref{alg:avo} finally outlines the full adversarial variational
optimization procedure, as built on top of WGAN-GP and NES.
Let us note however that the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

>>>>

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likehood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem by minimizing variationial upper bounds of the adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
\glnote{Mention experimental results.}
\glnote{Add 'so what?' conclusion.}

@@

In this note, we develop a likelihood-free inference algorithm for fitting a
forward non-differentiable generative model to observed data. We propose to
adapt the adversarial training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the
adversarial objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an abitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

\subsection{Variational optimization}

@@

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound $U(\psi)$ is differentiable, and using the log-likelihood trick it comes:
\begin{align}\label{eqn:approx-grad}
    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] \nonumber \\
    &= \nabla_\psi \int f(\mathbf{c})  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \int f(\mathbf{c}) \nabla_\psi q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \int \left[ f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c}) \right]  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c})]
\end{align}
Effectively, this means that provided that the score function $\nabla_\psi \log q_\psi(\mathbf{c})$ of the proposal
is known and that one can evaluate $f(\mathbf{\mathbf{c}})$ for any $\mathbf{c}$, then
one can construct empirical estimates of Eqn.~\ref{eqn:approx-grad}, which can
in turn be used to minimize $U(\psi)$ with stochastic gradient descent (or a variant thereof, like Adam~\cite{2014arXiv1412.6980K} to account for scaling invariance and robustness to noisy gradients).

% In this simple form, stochastic gradient descent based on
% Eqn~\ref{eqn:approx-grad} is known to suffer from premature convergence and lack
% of scale invariance. To avoid the dependence on the parameterization of the
% proposal distribution, natural gradients~\citep{amari1998natural} can be used
% instead to point towards the steepest descent in the space of realizable
% proposal distributions, rather than in the space of their parameters. Effectively,
% the natural evolution strategy (NES) algorithm~\citep{2011arXiv1106.4487W}
% corrects the gradient of $U$ according to the local curvature of the KL-divergence
% surface of proposal distributions, which results in forming
% \begin{equation}\label{eqn:approx-grad-nat}
%     \tilde{\nabla}_\psi U(\psi) = \mathbf{F}(\psi)^{-1} \nabla_\psi U(\psi),
% \end{equation}
% where $\mathbf{F}(\psi)$ is the Fisher information matrix
% \begin{equation}
%     \mathbf{F}(\psi) = \mathbb{E}_{\theta \sim q_\phi}[ \nabla_\psi \log q_\psi(\theta) \nabla_\psi \log q_\psi(\theta)^{\top} ]
% \end{equation}
% and can be estimated from samples, in particular by reusing the derivatives $\nabla_\psi \log q_\psi(\theta)$ already
% evaluated to form $\nabla_\psi U(\psi)$.

@@

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q_\psi \leftarrow \text{prior on $\theta$ (with differentiable and known likelihood)}$}
        \While{$\psi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \}_{m=1}^M \sim p_r$, $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$, $\{\epsilon_m \}_{m=1}^M \sim U[0,1]$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{Adam}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M U_d^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{\mathbf{z}_m \}_{m=1}^M \sim p_z$, $\{\theta_m \}_{m=1}^M \sim q_\psi$.}  \Comment{Update $q_\psi$}
            \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
            % \State{$\mathbf{F}(\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi \log q_\psi(\theta_m) \nabla_\psi \log q_\psi(\theta_m)^{\top}$}
            \State{$\nabla_\psi H(q_\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m)$}
            \State{$\psi \leftarrow \text{Adam}(\nabla_\psi U_g + \gamma \nabla_\psi H(q_\psi))$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

@@

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q_\psi(\theta)$ over the
parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$g$, estimates of $\nabla_\psi U_g$ and $\tilde{\nabla}_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
% That is,
% \begin{equation}
%     \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
% \end{equation} which we can approximate with mini-batches of
% generated data
% \begin{equation}
%     \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
% \end{equation}
% for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
% In practice, using
% \begin{equation}
%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
% \end{equation}
% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}

@@

For completeness, Algorithm~\ref{alg:avo} finally outlines the full adversarial variational
optimization procedure, as built on top of WGAN-GP.
Let us note however that the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.
